---
title: "Sardine Story"
author: "Charlotte Bolch"
date: "9/16/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to story here with motivation for the analysis

### Protagonist name- Janet

Pl. 1 
Shot of Janet at the computer in your bedroom getting ready to start the analysis of the sardine data. 
Janet: I'm really looking forward to diving into this data to see what I can figure out what happened to the sardine industry! 

Pl. 2
Zoom in on Janet's computer which has a sample of the sardine dataset. 
Janet: Well, I see in this dataset of ocean temperatures a lot of rows that have the same year. So this must mean that each row in this dataset has a recording of the ocean temperature for that day. 

Pl. 3
Zoom in on Janet's face that looks overwhelmed
Janet: Oh wow! That is a lot of data! I think I'm going to need to call on some help from Acme Data Consulting to figure out how to clean this data before I can start the analysis. 

Pl. 4
Split shot of Janet on the phone talking to DataDog
Janet: Hi DataDog! I'm in need of some consulting help figure out how to process this dataset of daily ocean temperatures. Can you help me?
DataDog: Did you say data? I love data! All about the data and the coding. We at Acme Data Consulting are all about programming perfection. 
Janet: Great! Let me tell you what my research questions are and show you the datasets that I have...

Pl. 5
Shot of Janet with word blurb for the following questions.

1. What is the relationship between the sardine catch and ocean temperature during peak spawning season?

2. How can I predict the sardine landings (total sardine catch) using ocean temperature data?

# Data Setup and Cleaning
## With DataDog

## Environment setup and load the two datasets.  
The first is ocean temperature readings from La Jolla, CA, the second is sardine landings (total catch) in tons.

```{r data setup}
library(dplyr)

temps_df    <- read.csv("~/UF/Fall 2019/stories/data/scripps_ocean_temps.csv")
landings_df <- read.csv("~/UF/Fall 2019/stories/data/sardine_landings_ueber.csv")

```

## Data cleaning
Because the onset of sardine spawning begins in February, create a subset of the temperature 
data from February to April.  This is the peak spawning season.  Then create a new data set 
of average temperatures for each year of the dataset. 
 
## Technical video about data manipulation HERE
 
```{r data cleaning}
temps_df <- temps_df[,-5]
temps_df <- na.omit(temps_df)
temps_df <- subset(temps_df, month == 2 | month == 3 | month == 4)
temps_df <- subset(temps_df, year < 1968)

avg_temps_df <- temps_df %>%
                  group_by(year) %>%
                  summarize(mean(surf_temp_c))

colnames(avg_temps_df)[2] <- "avg_temp"
```


## Graphing the data

Graph the raw temperature data

```{r graph temp}

plot(avg_temps_df$year, avg_temps_df$avg_temp,                  
     type = "l",
     col  = "blue",
     main = "",
     xlim = c(1915, 1970),
     xlab = "Year",
     ylab = 'Temperature (Celsius)')

```

### QUESTIONS:
1. What do you observe from this line graph? 
    i. Does the temperature of the ocean change over the years? If so, how?

2. Are you able to identify an overall trend from this line graph?


Graph the raw sardine landings data

```{r graph landings}

plot(landings_df$year, landings_df$monterey_tons,                  
     type = "l",
     col  = "blue",
     main = "",
     xlim = c(1915, 1970),
     xlab = 'Year',
     ylab = 'Landings (Tons)')

```

### QUESTIONS:
1. What do you observe from this line graph? 
    i. Does the amount of sardine landings change over the years? If so, how? 
    
    
2. Are you able to identify an overall trend from this line graph?

3. Looking at both line graphs of the raw temperature and sardine landings data, do you notice any relationship between the two variables? Take note of the x-axis which allows us to compare these two graphs. 

4. Is there a statistical technique that could be used to better see the trend in each of these variables?

## Story check-in here?

## Smoothing Technique: Loess Line

Let's try a smoothing technique called the loess line.
To do this we need to create a temperature model (loess), generate predicted values using that model, and then lay down a smoothing line on top of the graph.

## Technical video about the loess line HERE

```{r loess temp}

temp_model <- loess(avg_temp ~ year, data = avg_temps_df, span = .50)
temp_line  <- predict(temp_model)

plot(avg_temps_df$year, avg_temps_df$avg_temp,                  
     type = "l",
     col  = "blue",
     main = "",
     xlim = c(1915, 1970),
     xlab = "Year",
     ylab = 'Temperature (Celsius)')

lines(temp_line, x = avg_temps_df$year, col = 'red')

```

Try the same method you used for the temperature data for the sardine landings data. 

```{r loess landings}

landing_model <- loess(monterey_tons ~ year, data = landings_df, span = .50)
landing_line  <- predict(landing_model)

plot(landings_df$year, landings_df$monterey_tons,                  
     type = "l",
     col  = "blue",
     main = "",
     xlim = c(1915, 1970),
     xlab = 'Year',
     ylab = 'Landings (Tons)')

lines(landing_line, x = landings_df$year, col = 'red')

```

### QUESTIONS:
1. What do you observe from each of these line graphs with the loess curve? 
    i. Are you able to identify a better overall trend using the loess lines for the temperature and sardine landings variables?
    
2. Can you describe the relationship between the two variables better now with the loess lines?

3. What do you notice between the two line graphs? How do they compare? How are they different?

## Story check-in here?

Now that we can better identify the trend of the temperature and sardine landings over the years,
let's take a closer look at the relationship between these two variables. We know that sardines 
need an ideal temperature in order to catch and grow their population. In order to better understand
the change in sardine landings over the years, we will use the temperature data to predict the 
amount of sardine landings. This type of technique is called linear regression. 

## Technical video about simple linear regression and correlation HERE

## Raw Temperature and Landing Correlation and Regression

First compute a simple linear regression and Pearson's correlation on the raw data and then we'll do the same with the smoothed data and compare the two.
Data Note:  landings_df and temps_df contain raw data.  

Join the two datasets on year and retain only those years where the catch was greater than 500 tons.  When the catch was less than 500 tons, the fishery was effectively shutdown. And then graph the data.

## Merge and graph the raw data

```{r merge and graph raw data}

analysis_df_raw <- merge(avg_temps_df, landings_df, by = 'year')
analysis_df_raw <- subset(analysis_df_raw, monterey_tons > 500)
analysis_df_raw <- subset(analysis_df_raw, year < 1954)

plot(x = analysis_df_raw$avg_temp, y = analysis_df_raw$monterey_tons,
     xlab = 'Surface Temp (Celsius)',
     ylab = 'Tons')

```

### QUESTIONS:
1. What can you say visually about the relationship between the average surface temperatures and the catch in tons? 
    i. Is the relationship positive or negative? Strong or weak?
    
2. Think about where you would place the linear regression line on the graph. 

## Story check-in here?

## Correlation for the raw data

Calculate Pearson's r between the raw temperatures and the raw landings.

```{r corr raw}

raw_cor <- cor(analysis_df_raw$monterey_tons, analysis_df_raw$avg_temp)

```

### QUESTION:
1. Does the correlation value align with your visual interpretation of the relationship between the surface temperature and the sardine catch? 

## Simple linear regression with raw data

Generate a simple linear regression model and add the regression line to the graph. 

```{r slr raw}

fit_raw <- lm(analysis_df_raw$monterey_tons ~ analysis_df_raw$avg_temp)

plot(x = analysis_df_raw$avg_temp, y = analysis_df_raw$monterey_tons,
     xlab = 'Surface Temp (Celsius)',
     ylab = 'Tons')

abline(fit_raw, col = 'red')

```

### QUESTIONS: 
1. How does the linear regression line compare to your estimated regression line?

2. Overall, what does Pearson's r and the simple linear regression model tells us about surface temperature as a predictor for the sardine catch?

## Story Check-in here? Need a transition to the smoothed data

## Smoothed Temperature and Landing Regression & Correlation

Now that we've conducted a simple linear regression and correlation with the raw data, let's
do the same with the smoothed data from our loess model.  With the smoothing line vectors,
create two new dataframes.  Add a year variable, populate it with values from 1917 to
1955, and then join both into an analysis dataframe.  Remove rows from analysis_df for years
when the fishery was shut down.  Graph the new dataset.  

Data Note:  landing_df and temp_df contain smoothed data.  

## Merge and graph the smoothed data

```{r merge and graph smooth data}

landing_df <- data.frame("landings" = landing_line)
temp_df    <- data.frame("temps" = temp_line)

landing_df$year <- 1917:1967
temp_df$year    <- 1917:1967

analysis_df_smooth <- merge(temp_df, landing_df, by = 'year')
analysis_df_smooth <- subset(analysis_df_smooth, landings > 500)
analysis_df_smooth <- subset(analysis_df_smooth, year < 1954)

plot(x = analysis_df_smooth$temps, y = analysis_df_smooth$landings,
     xlab = 'Surface Temp (Celsius)',
     ylab = 'Tons')

```

### QUESTIONS: 
1. What can you say visually about the relationship between the smooth surface temperatures data and the smooth catch data? 
    i. Is the relationship positive or negative? Strong or weak?
    
2. Think about where you would place the linear regression line on the graph.

## Story check-in here?

## Correlation for the smoothed data

Calculate Pearson's r between the smooth temperature and landings data.

```{r corr smooth}

smooth_cor <- cor(analysis_df_smooth$landings, analysis_df_smooth$temps)

```

### QUESTION:
1. Does the correlation value align with your visual interpretation of the relationship between the surface temperature and the sardine catch? 

## Simple linear regression with smoothed data

Generate a simple linear regression model and add the regression line to the graph for the smooth data. 

```{r slr smooth}

fit_smooth <- lm(analysis_df_smooth$landings ~ analysis_df_smooth$temps)

plot(x = analysis_df_smooth$temps, y = analysis_df_smooth$landings,
     xlab = 'Surface Temp (Celsius)',
     ylab = 'Tons')

abline(fit_smooth, col = 'red')

```

### QUESTIONS: 
1. Does the correlation value for the smoothed data align with your visual interpretation of the 
relationship between the smoothed surface temperature data and the smoothed sardine catch data? 

2. How does the linear regression line of the smoothed data compare to your estimated regression line?

### Overall Questions
1. Overall, how does the linear regression line and the correlations compare between the smoothed
and raw data? 

2. What does this tell us about using temperature data to predict sardine landings?

# Exercise 

## Looking at data from 1970 to 2018
Now that we have looked at the sardine catch and ocean temperature data before 1968, let's investigate any possible trends in the data after 1968. 

#### What question do you want this exercise to be based on? 

### Steps:
1. Subset the sardine catch and temperature data to have year >= 196 and month only Feburary to April.

2. Create separate graphs from the raw data for temperature and sardine landings. 

3. QUESTIONS: 
    i. What do you see from the graphs? What comparisons can you make between the data from before 1968 and the data after 1968? 
    ii. What data should be used to move forward in this investigation? 
    
4. Use the temperature data to predict sardine landings after 1968. Make an appropriate graph with a linear regression line using the data selected and interpret the correlation value. 

5. QUESTION: 
    i. What do you see from the graph using the subset of data after 1968? 
    



