--- 
title: |
       | Activity Worksheet
       | *Loops*
author: "Dan Maxwell"
date: "June 19, 2017"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("gapminder.csv", header=TRUE, stringsAsFactors = FALSE)

```
## Groundhog Day
In the comedy *Groundhog Day*, Bill Murray plays Phil Connors, a disgruntled TV 
weatherman who's sent to Punxsutawney, PA to cover the story of Punxsutawney Phil, the
town's famous groundhog who can predict either an early spring or extended winter
depending on whether he sees his shadow or not.  While there, Phil Connors gets caught
in a time warp, reliving the same day over and over again until he learns to value 
each moment and the people he works with.

Here are a couple of *Groundhog Day* resources:

* [Movie Trailer](https://www.youtube.com/watch?v=2vmmTnDJnH0)
* [Groundhog Day Wikipedia](https://en.wikipedia.org/wiki/Punxsutawney_Phil)

In a programming language, loops allow you to repeat a specific action over and over again.  This grants you special powers as a programmer.  For example, you could define a groundhog loop that repeats a specified number of times, forcing Phil Connors to relive the same day until the loop reaches a defined limit.  The R code to accomplish this feat is listed below.  Try running this code in the console.

```{r eval=FALSE}

# Request user input -- the number of days to loop through.
ghog_days <- readline(prompt = "Please enter the number of groundhog days: ")

# Loop for the number of days specified in the ghog_days variable.
for(idx in 1:ghog_days) {
  msg <- paste("When will this day end? I've already lived it", idx, "times!")
  print (msg)
}

```

Many times our code begins to look like a bad sequel to *Groundhog Day*.  A piece of code works well so we copy it with minor modifications and then do it again and again, as in the case below. In this example, the code that creates the life expectancy plot is repeated three times, once for each country; the key difference between each being the subset of data that is graphed.

Although this code works and correctly produces three graphs, it is not optimal.  In fact, it breaks one of the cardinal rules of good programming: 

>> **Thou Shalt Not Repeat Thyself**

Rewriting the same code with minor modifications is similar to Phil Connor's experience of living groundhog day over and over.  At some point, the underlying repetition becomes unbearable.  And, in the case of code, it is unsustainable.  Imagine the maintenance nightmare if we repeated this same `plot()` code with minor modifications 25 times.  In that situation, you'd need to change each block of code for a total of 25 updates.

\newpage

```{r fig.width = 5, fig.height = 10, eval=FALSE}

df_tmp <- read.csv("gapminder.csv", header = TRUE, stringsAsFactors = FALSE)

opar <- par(no.readonly = TRUE)
par(mfrow = c(3, 1))

df_gap <- df_tmp[df_tmp$country == 'United States',]

plot(df_gap$year, df_gap$lifeExp, main = "Life Expectancy (United States)", 
     xlab = "Year", ylab = "Years", col = "blue", type = "l", xlim = c(1950,2010))


df_gap <- df_tmp[df_tmp$country == 'France',]

plot(df_gap$year, df_gap$lifeExp, main = "Life Expectancy (France)", 
     xlab = "Year", ylab = "Years", col = "blue", type = "l", xlim = c(1950, 2010))


df_gap <- df_tmp[df_tmp$country == 'Brazil',]

plot(df_gap$year, df_gap$lifeExp, main = "Life Expectancy (Brazil)", 
     xlab = "Year", ylab = "Years", col = "blue", type = "l", xlim = c(1950, 2010))

# Reset original graphing parameters
par(opar) 

```

So how do we put a stop to this madness?  Well, maybe we can use the idea behind our original groundhog loop to solve this problem.  Although the experience wasn't fun for Phil Connors, it might prove useful here.

The first step is to carefully compare the lines of code above to determine what changes in each sub-section.  Immediately, we see that the each subsets the data from the df_tmp dataframe differently.  The country name changes.  The title of each plot also changes (passed as an argument to main), depending on the country being graphed. It looks like country_name is the variable of interest here.

> ## Question
>
> Question text goes here...

> ## Challenge 
>
> Challenge text goes here...
>
> > ## Solution 
> > Solution text goes here...
> >
> > ```{r sample_code, eval=FALSE}
> > rows2002_number <- nrow(gapminder[(gapminder$year == 2002),])
> > ```
> >

## Repeating operations

If you want to iterate over a set of values, when the order of iteration is important, and perform the same operation on each, a `for()` loop will do the job.  This is the most flexible of looping operations, but therefore also the hardest to use correctly. Avoid using `for()` loops unless the order of iteration is important: i.e. the calculation at each iteration depends on the results of previous iterations.

The basic structure of a `for()` loop is:

```{r, eval=FALSE}
for(iterator in set of values){
  do a thing
}
```

For example:

```{r}
for(i in 1:10){
  print(i)
}
```

The `1:10` bit creates a vector on the fly; you can iterate
over any other vector as well.

We can use a `for()` loop nested within another `for()` loop to iterate over two things at
once.

```{r}
for(i in 1:5){
  for(j in c('a', 'b', 'c', 'd', 'e')){
    print(paste(i,j))
  }
}
```

Rather than printing the results, we could write the loop output to a new object.

```{r}
output_vector <- c()
for(i in 1:5){
  for(j in c('a', 'b', 'c', 'd', 'e')){
    temp_output <- paste(i, j)
    output_vector <- c(output_vector, temp_output)
  }
}
output_vector
```

This approach can be useful, but 'growing your results' (building
the result object incrementally) is computationally inefficient, so avoid
it when you are iterating through a lot of values.


> ## Tip: don't grow your results
>
> One of the biggest things that trips up novices and
> experienced R users alike, is building a results object
> (vector, list, matrix, data frame) as your for loop progresses.
> Computers are very bad at handling this, so your calculations
> can very quickly slow to a crawl. It's much better to define
> an empty results object before hand of the appropriate dimensions.
> So if you know the end result will be stored in a matrix like above,
> create an empty matrix with 5 rows and 5 columns, then at each iteration
> store the results in the appropriate location.

A better way is to define your (empty) output object before filling in the values.
For this example, it looks more involved, but is still more efficient.

```{r}
output_matrix <- matrix(nrow=5, ncol=5)
j_vector <- c('a', 'b', 'c', 'd', 'e')
for(i in 1:5){
  for(j in 1:5){
    temp_j_value <- j_vector[j]
    temp_output <- paste(i, temp_j_value)
    output_matrix[i, j] <- temp_output
  }
}
output_vector2 <- as.vector(output_matrix)
output_vector2
```

> ## Tip: While loops
>
>
> Sometimes you will find yourself needing to repeat an operation until a certain
> condition is met. You can do this with a `while()` loop.
>
> ```{r, eval=FALSE}
> while(this condition is true){
>   do a thing
> }
> ```
>
> As an example, here's a while loop
> that generates random numbers from a uniform distribution (the `runif()` function)
> between 0 and 1 until it gets one that's less than 0.1.
>
> ~~~
> z <- 1
> while(z > 0.1){
>   z <- runif(1)
>   print(z)
> }
> ~~~
> {: .r}
>
> `while()` loops will not always be appropriate. You have to be particularly careful
> that you don't end up in an infinite loop because your condition is never met.

> ## Challenge 3
>
> Write a script that loops through the `gapminder` data by continent and prints out
> whether the mean life expectancy is smaller or larger than 50
> years.
>
> > ## Solution to Challenge 3
> >
> > **Step 1**:  We want to make sure we can extract all the unique values of the continent vector
> > ```{r 07-chall-03-sol-a, eval=FALSE}
> > gapminder <- read.csv("data/gapminder-FiveYearData.csv")
> > unique(gapminder$continent)
> > ```
> >
> > **Step 2**: We also need to loop over each of these continents and calculate the average life expectancy for each `subset` of data.
> > We can do that as follows:
> >
> > 1. Loop over each of the unique values of 'continent'
> > 2. For each value of continent, create a temporary variable storing the life exepectancy for that subset,
> > 3. Return the calculated life expectancy to the user by printing the output:
> >
> > ```{r 07-chall-03-sol-b, eval=FALSE}
> > for( iContinent in unique(gapminder$continent) ){
> >    tmp <- mean(subset(gapminder, continent==iContinent)$lifeExp)
> >    cat("Average Life Expectancy in", iContinent, "is", tmp, "\n")
> >    rm(tmp)
}
> > ```
> >
> > **Step 3**: The exercise only wants the output printed if the average life expectancy is less than 50 or greater than 50. So we need to add an `if` condition before printing.
> > So we need to add an `if` condition before printing, which evaluates whether the calculated average life expectancy is above or below a threshold, and print an output conditional on the result.
> > We need to amend (3) from above:
> >
> > 3a. If the calculated life expectancy is less than some threshold (50 years), return the continent and a statement that life expectancy is less than threshold, otherwise return the continent and   a statement that life expectancy is greater than threshold,:
> >
> > ```{r 07-chall-03-sol-c, eval=FALSE}
> > thresholdValue <- 50
> >
> > for( iContinent in unique(gapminder$continent) ){
> >    tmp <- mean(subset(gapminder, continent==iContinent)$lifeExp)
> >    
> >    if(tmp < thresholdValue){
> >        cat("Average Life Expectancy in", iContinent, "is less than", thresholdValue, "\n")
> >    }
> >    else{
> >        cat("Average Life Expectancy in", iContinent, "is greater than", thresholdValue, "\n")
> >         } # end if else condition
> >    rm(tmp)
> >    } # end for loop
> >
> > ```

> ## Challenge 4
>
> Modify the script from Challenge 4 to loop over each
> country. This time print out whether the life expectancy is
> smaller than 50, between 50 and 70, or greater than 70.
>
> > ## Solution to Challenge 4
> >  We modify our solution to Challenge 3 by now adding two thresholds, `lowerThreshold` and `upperThreshold` and extending our if-else statements:
> >
> > ```{r 07-chall-04-sol, eval=FALSE}
> >  lowerThreshold <- 50
> >  upperThreshold <- 70
> >  
> > for( iCountry in unique(gapminder$country) ){
> >     tmp <- mean(subset(gapminder, country==iCountry)$lifeExp)
> >     
> >     if(tmp < lowerThreshold){
> >         cat("Average Life Expectancy in", iCountry, "is less than", lowerThreshold, "\n")
> >     }
> >     else if(tmp > lowerThreshold && tmp < upperThreshold){
> >         cat("Average Life Expectancy in", iCountry, "is between", lowerThreshold, "and", upperThreshold, "\n")
> >     }
> >     else{
> >         cat("Average Life Expectancy in", iCountry, "is greater than", upperThreshold, "\n")
> >     }
> >     rm(tmp)
> > }
> >  ```

> ## Challenge 5 - Advanced (Optional)
>
> Write a script that loops over each country in the `gapminder` dataset,
> tests whether the country starts with a 'B', and graphs life expectancy
> against time as a line graph if the mean life expectancy is under 50 years.
>
> > Solution for Challenge 5
> >
> > We will use the `grep` command that was introduced in the Unix Shell lesson to find countries that start with "B."
> > Lets understand how to do this first.
> > Following from the Unix shell section we may be tempted to try the following
> > ```{r 07-chall-05-sol-a, eval=FALSE}
> > grep("^B", unique(gapminder$country))
> > ```
> >
> > But when we evaluate this command it returns the indices of the factor variable `country` that start with "B."
> > To get the values, we must add the `value=TRUE` option to the `grep` command:
> >
> > ```{r 07-chall-05-sol-b, eval=FALSE}
> > grep("^B", unique(gapminder$country), value=TRUE)
> > ```
> >
> > We will now store these countries in a variable called candidateCountries, and then loop over each entry in the variable.
> > Inside the loop, we evaluate the average life expectancy for each country, and if the average life expectancy is less than 50 we
> > use base-plot to plot the evolution of average life expectancy:
> >
> > ```{r 07-chall-05-sol-c, eval=FALSE}
> > candidateCountries <- grep("^B", unique(gapminder$country), value=TRUE)
> >
> > for( iCountry in candidateCountries){
> >     tmp <- mean(subset(gapminder, country==iCountry)$lifeExp)
> >     
> >     if(tmp < thresholdValue){
> >         cat("Average Life Expectancy in", iCountry, "is less than", thresholdValue, "plotting life expectancy graph... \n")
> >         
> >         with(subset(gapminder, country==iCountry),
> >                 plot(year,lifeExp,
> >                      type="o",
> >                      main = paste("Life Expectancy in", iCountry, "over time"),
> >                      ylab = "Life Expectancy",
> >                      xlab = "Year"
> >                    ) # end plot
> >               ) # end with
> >     } # end for loop
> >     rm(tmp)
> >  }
> >  ```

### References
Ekstrom, C. T. (2017). *The R primer (2nd ed.)*. Boca Raton: CRC Press.

Ott, L., & Longnecker, M. (2016). *An introduction to statistical methods & data analysis* (7th ed.). Boston, MA: Cengage Learning.

