---
title: "bird_story_analysis"
author: "Charlotte Bolch"
date: "3/11/2020"
output: 
  slidy_presentation:
    font_adjustment: -1
    footer: "Copyright 2019 University of Florida Research Foundation, Inc. All Rights Reserved."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Objective
1. Students will be able to demonstrate their knowledge of the process of data collection.
2. Students will be able to create visualizations to look at the relationship among the variables of species richness, elevation, and habitat types. 

## Panels 1-12 (Google Slides)

## Data Learning Event (Panels 13 to 16)

Before Statcat goes out into the field to collect the data, he would like to look at the data that has already been collected by the cats. Statcat has the data provided to him and he is interested in looking at the different habitat types. He first needs to load the relevant packages and read the data into R. 

### Environment setup and loading datasets  

```{r data setup, warning=FALSE, message=FALSE}

# Load packages (make sure all packages are installed first)
library(dplyr) #coding structure function
library(car) #Scatterplot function

#setwd("C:/stories/_bird")

# Read in data set collected from DataDog
cat_species_df <- read.csv("../data/cat_species_rich.csv")

```

### Habitat type bar graph
To look at all of the different habitat types, we will create a bar graph. We first need to create a table of the different types of habitats and then graph that table using the function barplot(). 

```{r habitat bar graph, warning=FALSE, message=FALSE}

#table of habitat types
counts_habitat <- table(cat_species_df$habitat)
counts_habitat

#bar graph of table
barplot(counts_habitat,
        main = "Counts of Habitat Types",
        xlab = "Habitat Types",
        ylab = "Count",
        ylim = c(0,30))

```

**Questions:** What is the story that is going on in this bar graph?

## Panels 17-24

## Data Learning Event (Panels 25-27)
After talking with the cats, StatCat has learned that the research questions that they are interested in are:

### Research Questions

1. What is the relationship between species richness and habitat types (forest, shrubs, trees, etc.)?
2. What is the relationship between area and habitat types?
3. What is the relationship between species richness and area for all bird communities?
4. What is the relationship between species richness and area for specific habitats of interest?

StatCat is on his way to start locating data from random locations using their GPS systems in order to answer those questions. The setup of the data collection form is important to plan out before StatCat starts collecting data. This way StatCat collects and records all the information that is needed from each location. The information that needs to be collected from each location is:
1. Site
2. Block
3. Area of habitat (Ha)
4. Richness
5. Habitat type
**An example of a data collection form is here**

## Panels 28-30

## Data Learning Event (Panels 31-35)

Using the data from the cats, let's look at species richness by habitat type (research question 1) to see if what the birds and DataDog are talking about is correct. What does the relationship between species richness and forest habitat look like compared to the other habitat types?

```{r richness boxplot, warning=FALSE, message=FALSE}

#variable = rich for species richness

#boxplot of rich by habitat
boxplot(rich ~ habitat, data=cat_species_df,
        main = "Species Richness by Habitat Type",
        xlab = "Habitat Type",
        ylab = "Species Richness")

```

When looking at a boxplot, remember that the solid black line is the the median value, the box indicates the values for the first and third quartiles, and the whiskers with the dotted lines indicate the minimum and maximum values. 

**Question:** What is the story that is going on in this set of boxplots?

**Alert!** Statcat has just uploaded the data that he has collected out in the field to the shared drive. Let's take a look at his data to investigate area among all habitat types (research question 2).

**Student exercise: Boxplot with Statcat data**

Create a boxplot for area by habitat
```{r boxplot student exercise, warning=FALSE, message=FALSE}

#Example code for students
# boxplot( ~ , data=,
#          main = "",
#          xlab = "",
#          ylab = "")

#Answer
boxplot(ha ~ habitat, data=cat_species_df,
         main = "Area by Habitat Type",
         xlab = "Habitat Type",
         ylab = "Area")

```

**Reflection:** What is the story that is going on in this boxplot? 

## Panels 36-45

## Data Learning Event (Panels 46-50)

StatCat and DataDog are interested in looking at the relationship between species richness and area for all habitat communities (research question 3). 

```{r rich area all scatterplot, warning=FALSE, message=FALSE}

scatterplot(rich~ha,data=cat_species_df,smooth=F)

```

DataDog: We can't see anything, let me take the log of Ha. 

```{r rich area all scatterplot log, warning=FALSE, message=FALSE}

scatterplot(rich~ha,data=cat_species_df,smooth=F,log="x",
            xlab = "log(Ha)")
title("Graph of All Habitats", adj = .5)

```

**Reflection:** What is the story that is going on in this scatterplot?

Statcat: Let's try looking at a similar graph but with only fragment habitats from my data (research question 4). Why don't you give it a try consultant?

Consultant, you first need create a subset of the data from StatCat for only habitat types that are "fragments". Then, they create a scatterplot of richness by log(area). 

# Richness with fragment bird communities graph
```{r rich area all scatterplot log, warning=FALSE, message=FALSE}

fragment_df <- filter(cat_species_df, habitat == "fragment") #creating a subset of fragment habitat only

scatterplot(rich~ha,data=fragment_df,smooth=F,log="x",
            xlab = "log(Ha)")
title("Graph of Only Fragment Habitats", adj = .5)

```

**Reflection:** What is the story that is going on in this scatterplot?

## Panels 51-54
### story ends


