---
title: "bird_story_analysis"
author: "Charlotte Bolch"
date: "2/13/2020"
output: 
  slidy_presentation:
    font_adjustment: -1
    footer: "Copyright 2019 University of Florida Research Foundation, Inc. All Rights Reserved."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Research Questions

1. What is the relationship between species richness and elevation for all bird communities?
2. What is the relationship between species richness and elevation by habitat types (forest, shrubs, trees, etc.)?
3. How can species richness be predicted by elevation within forest habitat?

## Environment setup and loading datasets  

```{r data setup, warning=FALSE, message=FALSE}

# Load packages (make sure all packages are installed first)
library(dplyr) #coding structure function
library(NCStats) #Subset function
library(vegan) #Ordinations
library(car) #Scatterplot function
library(sciplot) #Bargraph.CI function

setwd("C:/stories/_bird")

# Read in bird data enironment
bird_data <- load(file = '../data/bird_data.RData') 

```

# Richness graphs
```{r richness graph, warning=FALSE, message=FALSE}

#relabel dataset
bird_df <- all

#factor levels for habitat variable
bird_df$habitat <- factor(hab$habitat,levels=c("forest","shrubs","fragment","tree","fence","pasture"))

#boxplot
boxplot(rich ~ habitat, data=bird_df) # ADD title and relabel axes
bargraph.CI(habitat,rich,data=bird_df) #adjust scale

```

Questions: What is the story that is going on in this boxplot? What is the story that is going on in this bar graph?

# Richness with forest bird communities graph
```{r richness forest bird, warning=FALSE, message=FALSE}

# Richness decreases with elevation for forest bird communities
forest <- Subset(all,habitat=="forest")
scatterplot(rich ~ elevation, data=forest, smooth=F)

```

# Model: Richness and elevation for forest bird communities
```{r richness and elevation model, warning=FALSE, message=FALSE}

mod.elev <- lm(rich~elevation,data=forest)
summary(mod.elev)

```

