---
title: "Fountain of Youth/Phsodex"
output: html_document
---

# ------------------------------------------------------------------
# MyDataStory - Fountain of Youth?
# By Scott Cohen
# ------------------------------------------------------------------

*Core Concepts:*
1. Introduction to Handling Datasets
2. Developing Analysis Plan 
3. Correlation

*Chapters:*
1. Introduction
2. Importing/Understanding
3. Merging
4. Visualize
5. Correlation
6. Exercise

*Learning Objectives:*
1. Importing and Understanding Dataframes in R
2. Developing an Analysis Plan
3. Joining/Merging Dataframes in R
4. Visualize Relationships in R
5. Hypothesis Testing - Correlation in R
6. Foundational Module Exercise

*Characters:*
**StatCat**
Core Traits - Old-fashioned, knows the statistics concepts very well, does not trust/afraid of programming
Story-Specific Traits - Manic behavior, obsessed with proving the story of his family, constantly questioning DataDogs methods
**DataDog:**
Core Traits - Modernized, short-attention span, scratches the surfaces with statistics, just knows how to do things through programming
Story-Specific Traits - Frustrated with StatCat's old-fashioned ways, 

# ------------------------------------------------------------------
# Chapter 1 - Introduction
# ------------------------------------------------------------------

Pl. 1
*SFX: Clunky computer sounds* 
*StatCat pulls a big printout out of his antiquated UNIVAC 1 computer.*

Pl. 2
*StatCat takes out a big pair of scissors and begins to cut the printout into pieces to subset his dataset, DataDog watches in confusion, StatCat has a very messy workspace*
**DataDog:** What...What are you doing?

Pl. 3
*StatCat focused on the dataset, mumbles to himself quietly*
**StatCat:** It has to be a real...The Fountain of Youth just has to be real...

Pl. 4
*Shot of DataDog's looking at StatCat in confusion, StatCats eyes remain focused on the printout*
**DataDog:** Fountain of Youth?

Pl. 5
*Shot of StatCat turning to face DataDog, eagerly explaining*
**StatCat:** My plan is simple yet beautiful. I just read about the Fountain of Youth in Florida and I am going to prove its truths! Illudium Phosdex! Its there!
**StatCat:** There is this story has been passed down from generation to generation! My grandfather used to tell me of how this mysterious chemical, Illudium Phosdex, can give cats the ability to live another 200 years! And this is in addition of my 9 lives!

Pl. 6
*DataDog starring at StatCat "blankly", StatCat holding up a newspaper clipping*
**StatCat:**: Look! This article says that there was high levels of Illudium Phosdex coming from the famous Fountain of Youth! I must use my skills to defend my family's name and honor...and live forever of course!
         
Pl. 6
*DataDog confused*
**DataDog:** Illudi...what? Oh forget it, good luck with all that.

Pl. 7
*Shot of DataDog walking away in disinterest, StatCat continues to mumble to himself, fumbling with his data sheets*
**StatCat:** Okay now to get started...step one is to cut up this printout and tape back together the peices I need

Pl. 7
*DataDog stops in his tracks, eyes open up wide, turns towards StatCat*
**DataDog:** Cut?...tape?...You live in the stone age!  I can write code that will do that in milliseconds...
**StatCat:** Oh really, my little furry friend.  I triple dog dare ya to do it...

Pl. 8
*DataDog stares at the viewer wide-eyed...*
**DataDog (thought):** There are times in a dog's life when she must stand for truth, justice, and the canine way...

Pl. 9
*DataDog with super cape and paws holding a light saber in preparation for battle with the evil rubber chicken empire.*
*DataDog reading books in her giant programming library.*
*SFX: Trumpet sound*
**DataDog (thought):** Let's do this...

# ------------------------------------------------------------------
# Chapter 2:
# Goal: Use base R functions to import and summarize the contents of the Florida Census 2010 dataframe 
# Learning Objective: Importing and Understanding Dataframes in R
# ------------------------------------------------------------------
*Sets scene showing DataDog busy, hacking into UNIVAC, digging threw stacks of paper, finding random hairballs, clunking sounds* 

Pl.
*Add joke about the mess*
**DataDog:** Why do we even bother keeping this junk laying around? 

Pl.
*DataDog Pulls out a two tape rolls, reads "TOP SECRET"*

Pl.
*StatCat jumps in*
**StatCat** Hey! Be careful with all of that! That tape you got there is very important to my family. I will have to give you access to that. I knew you were going to need my help eventually... wait.. what are you doing?

Pl.
*DataDog nonchalantly loads the tape into his computer, without the help of StatCat*
**DataDog** You always underestimate my skills, I'll show you how its done. Look I have one of the US Census file loaded into my computer already. All I needed to do is define the location of the file, called a working directory, with the the setwd() function then import the file with the read.csv() function.

```{r}
#set working directory, this 'tells' R where the file path of where the data is located.
setwd("/Users/ScottCohen/OneDrive - University of Florida/MyDataStory/")

#import data, naming the dataframe as census_df
census_df<-read.csv("data/census.csv",stringsAsFactors = F) 
```

Pl.
*StatCat interupts, DataDog continues to ignore*
**StatCat** Well, you are going to need to me to tell you what's in those files...

Pl.
*DataDog annoyed, throws a cat toy out of the panel*
**DataDog** No, I know how to do that too. Now, will you quit distracting me? Here, go find something else to do for now.

Pl.
*DataDog discussing with herself*
**DataDog:** Finally, let's dig into this census file some more. The dim(), names(), and str() functions should do the trick. These will all provide me information about the structure of this ol' cats files.

```{r}
#Dimensions of dataset, 4182 rows (observations) and 20 columns (variables)
dim(census_df)

#Names of the variables (columns)
names(census_df) 

#Summary of the number of observations, variables types, and first 5 values
str(census_df) 
```

Pl.
*Datadog turns back towards computer*
**DataDog:** So what does the data actully look like? I just need to print the data. I can enter the dataframe name by itself in order to print the entire dataset.

```{r}
#Full display
census_df

#Spreadsheet view
View(census_df)
```

Pl.
**DataDog:** Woah! Over 4000 observations! Too much! Let me just see first few observations? I can use the head() function to find the first 6 observations!
```{r}
#Default at 6 rows
head(census_df) 

#Use the "n" to change to only 4 rows
head(census_df, n=4) 
```

Pl.
*DataDog turns her shoulder, sees StatCat chasing his tail around*
**DataDog:** ...and the tail() function for the last 6 observations.

```{r}
#See the last 6 rows
tail(census_df)
```


**EXERCISE - Import and Summarize the Illudium Phosdex environmental file**

Pl.
*DataDog look towards the pile of StatCats old tapes*
**DataDog:** Now where was the other datafile that StatCat was working with...
**DataDog:** Ah! Here it is! It looks like its environmental sampling data from 2018. It says here in the file description that the phosdex.csv file contains the concentration measurements for the census tracts in the other file. We will just need to repeat the same steps for this file. 

Pl.
*Loud bang/noise coming from the panel*
*DataDog sees StatCat knocking over his box of bones*

Pl.
*DataDog annoyed*
**DataDog**: This cat is always knocking my stuff over! I'm going to have to get one of the new interns import this phosdex data while I'll deal with this mess.

*SOUND EFFECT 'CALLING' THE INTERN (STUDENT)*
*Need to integrate the student into the story, find a way to call the student by name to get them into the story*
*LOOK UP WAY TO DO THIS*

#Exercise
(1) Set working directory using setwd()
(2) Import data using read.csv() and name the dataframe as phosdex_df
(3) Summarize the structure of data using str() 
(4) Visualize the data using View()

```{r}
#Set working directory
setwd("/Users/ScottCohen/OneDrive - University of Florida/MyDataStory/")

#Use read.csv function
phosdex_df<-read.csv("data/phosdex.csv",stringsAsFactors = F) 
str(phosdex_df)
View(phosdex_df)
```


# ------------------------------------------------------------------
# Chapter 3
#
# Goal: Develop an analysis plan to assess the relationship between Median Age, Illudium Phosdex measurements, and distance from the the Fountain of Youth
# Learning Objective: Developing an Analysis Plan
# ------------------------------------------------------------------

Pl.
*StatCat bugging DataDog, asking for updates, Datadog slighlty annoyed*
**StatCat** What's going on over there?? Hurry up will ya? This cat is aging by the second!

Pl.
*DataDog shooing away StatCat*
**DataDog:** This would have taken you hours to complete! Theres is over 4,000 census tracts in this Census dataset and 442 measurements in the Illudium Phosdex data. I would have been done by now if you would just leave me alone. Go find somewhere to chill...

Pl.
*StatCat walking away; DataDog slightly frustrated, shaking her head*
**DataDog:** That cat has no idea how much time I am saving him... But I really need to develop a plan to analyze this data

Pl.
*StatCat overhearing DataDog*
**StatCat:** Did I hear "plan"?? Oh my furry friend, I knew you would need my help eventually. Sit down and let me tell you a little story.

Pl.
*NEED TO CREATE A NARRATIVE FOR INTRODUCING ANALYSIS PLAN*
*StatCat and DataDog steping over to a chalk/white board* 
**StatCat:** There are 5 general steps in any analysis plan. 
1. Define your research questions
2. Identify the datasets to be used
3. Set inclusion/exclusion criteria 
4. Select appropriate statistical tests
5. Interpret results
*StatCat presenting the newspaper clipping in one hand, picture of his family in the other*
*Moving pictures, cat meme idea to find a metaphor for each variable*
**StatCat:** Our research questions are two part. We must first replicate the news article and confirm the relationship between the Fountain of Youth and the Illudium Phosdex. Then we will assess the relationship between Illudium Phosdex and Age to prove my family's story!
1a. Is there a relationship between Illudium Phosdex concentration in local springs and distance from the Fountain of Youth?
1b. Is there a relationship between Illudium Phosdex concentration and Age?
*StatCat presenting the metadata, pop-up(?) frame that presents the list of variables in each dataset side by side, diagram connecting the variables of interests*
**StatCat:** Now onto identifying the datasets we are using! My family has been collecting Illudium Phosdex measurements and Census data across the country for years! These two datasets would be perfect for us to test our research questions. The Census data you have contains demographics of the census tracts in the state of Florida. Each census tract is given a unique ID, these are defined by the "GEOID" variable in the dataset. Now, before you took my data from me, I already calcuated the distance of each census tract from the Fountain of Youth. This is defined by the "distance" variable. The median age of each census tract is defined by the "age" variable.
**StatCat:**The measurements in the Illudium Phosdex dataset are randomly sampled across the state of Florida by the best equipment my family possess! We sample each census tract once throughout the year. Just like the other dataset, the same "GEOID" variable is used as the unique identifier for each census tract. Illudium Phosdex concentration levels are defined by the "phosdex" variable.

Pl.
*DataDog eager to get started*
**DataDog:** Sounds like we need to merge those two datasets into one! That should be easy...let me get started!
**StatCat:** Wait! We haven't finished the analysis plan yet!

Pl.
**StatCat:** In this newspaper article, it says Illudium Phosdex has been found up to 150 miles from the Fountain of Youth. We can set our inclusion criteria in the Census dataset to only include census tracts that are within 200 miles from the Fountain of Youth.
**DataDog:** Okay great! Sounds like we can subset...can we get started now?

Pl.
**StatCat:** You data folks are always so eager to get started without a full plan! You will run into trouble and waste time if you keep this up! 
**StatCat:** For our statistical analysis, we are using only continuous variables! So a simple correlation test should do the trick! Then we can interpret our results apporpriately based on the correlation coefficient.

Pl.
*DataDog looking at StatCat eyes open, blankly, waiting to get started*
**DataDog:** Soo....
**StatCat:** Yes! Get started already! I'll be packing my bags for our trip to the Fountain of Youth...

Pl.
*DataDog talking to herself*
**DataDog:** Now that I have my new census data, I am ready to merge it with the Illudium Phosdex.. This will create one complete dataset!

# ------------------------------------------------------------------
# Chapter 4
#
# Learning Objective: Joining/Merging Datasets in R
# Goal: Combine the new subset Florida Census and Illudium Phosdex dataframes into one
# ------------------------------------------------------------------
*Set scene with DataDog working, StatCat moving in the background "packing" (e.g. toys, cat-related objects), stops distracted by the TV*
*COMMERCIAL BREAK*
Narrator: Introducing the ...
Narrator: Now back to your regular scheduled programming

Pl. 25
**DataDog:** Since the Illudium Phosdex file contains a random sample of 250 census tracts, but we only want those that are less than 200 miles from the Fountain of Youth, we will use a natural (i.e. inner) join to link the Illudium Phosdex data frame and the census subset dataframe using the merge() function. GEOID is going to be the variable we basing the matching on, this is because it is the unique identifier in both datasets.


```{r}
#merge(x, y, by = ) = merge(Data Frame A, Data Frame B, Variable to base join)
phosdex_census_df <- merge(census_df, phosdex_df[ , c("geoid","date_sampled","phosdex")], 
                   by = "geoid")
head(phosdex_census_df)
```

# ------------------------------------------------------------------
# Chapter 5
#
# Learning Objective: Visualize Distributions and Relationships in R
# Goal: Illustrate the distribution and relationships using histrograms and scatterplots
# ------------------------------------------------------------------

Pl 27.
**DataDog:** Now we need to visualize the data using a scatterplot as a way to illustrate a relationship between two varies. This can be quickly done by create a scatterplot using the plot() function.

```{r}
plot(phosdex_census_df$distance, phosdex_census_df$phosdex, 
     main="Illudium Phosdex Concentration vs. Distance From Fountain of Youth for Each Florida Census Tract", 
     xlab="Distance", 
     ylab="Phosdex Concentration"
     )

plot(phosdex_census_df$age, phosdex_census_df$phosdex, 
     main="Median Age vs. Illudium Phosdex Concentration for Each Florida Census Tract", 
     xlab="Median Age", 
     ylab="Phosdex Concentration"
     )
```

*SOUND EFFECT 'CALLING' THE INTERN (STUDENT)*
*Need to integrate the student into the story, find a way to call the student by name to get them into the story*
*LOOK UP WAY TO DO THIS*

**EXERCISE**
(1) Create a scatterplot to illustrate the relationship between Illudium Phosdex concentration (phosdex) and median age (age)
(2) Thoughts on patterns

# ------------------------------------------------------------------
# Chapter 6
#
# Learning Objective: Hypothesis testing using Pearson's correlation
# Goal: Assess if there is a statistical association between age, illudium phosdex, and distance
# ------------------------------------------------------------------

Pl.
*DataDog pulls up the first scatterplot on the screen*
**DataDog:** Here kitty, kitty...  I've got something to show you.
**StatCat:** Hey - who you calling kitty?

Pl. 
*DataDog explaining results to StatCat, StatCat has his bags packed already*
**DataDog:** Look! It looks like there is some pattern with your Illudium Phosdex measurements and distance from the Fountain of Youth! This confirms the result from that newspaper article you have!

Pl.
*DataDog pulls up the second, second scatterplot*
**DataDog:** But... the other one...

Pl.
*StatCat shocked, drops his bags*
**StatCat:** No, no, no...there must be something there. I know! Let's statistically test to see if there is an association between Illudium Phosdex vs. Distance and Illudium Phosdex vs. Age. Come with me to my office! I'll show you how to do it!

Pl.
*StatCat frantically drags DataDog out of her chair, StatCat to focused to hear DataDog*
**DataDog:** But... I already know how...

Pl.
*StatCat's office is filled with pictures of "perfect correlations"*



```{r}
# cor.test(x, y, ...) = tests for association between paired samples, using Pearson's correlation or Spearmans's correlation coeffecients
cor.test(phosdex_census_df$distance, phosdex_census_df$phosdex)
cor.test(phosdex_census_df$age, phosdex_census_df$phosdex)
```

##EXERCISE
(2) Assess the correlation between Illudium Phosdex and age

Pl.
*StatCat waiting on DataDog to obtain the results*
**StatCat:** Well...whats the correlation coefficient?? There is an association right? Oh wow... I cannot wait to live forever!

Pl.
*DataDog turns towards StatCat*
**DataDog:** Hold on there, I know you are the expert with statistics...but there doesn't appear to be any correlation in the data with age and the Fountain of Youth

Pl.
*StatCat in denial*
**StatCat:** That's impossible! There has to be a mistake with your code! This Fountain of Youth story has been passed down from generation to generation in my family. 

Pl.
*DataDog insulted by StatCat*
**DataDog:** My code works perfectly!

Pl.
*StatCat disgruntled, storms off in the distance*
**StatCat:** Well I am telling you this story is true! This must be the wrong data. There has to be another dataset somewhere with the right age measurements...

Pl.
*StatCat storms off down a dark hallway*
*Text on panel, To be continued...*

# ------------------------------------------------------------------
# Chapter 6
# 
# Learning Objevtice: Foundational Module Final Exercise 
# Goal: Use the skills and functions learned in this module
# ------------------------------------------------------------------

Pl.
*StatCat in a dark basement, digging through dusty tape racks, *
**StatCat:** I must defend my family's honor, I know there is an old reference down here somewhere.

Pl.
*StatCat continues to running UNIVAC systems in search for code, eyes widen when he finds what he is looking, finds the tape with the "age" data*

Pl.
*StatCat excitedly approaches DataDog*
**StatCat:** Here! Found the right data file with the age measurements! DataDog, can you run your code using this?

Pl.
**DataDog:** What is it?
**StatCat:** Oh...it's just an older reference

**Exercise**
1. Import the new dataset called statcat_age.csv
2. Merge the distances from the FLCensus2010 to the new dataset
3. Subset the census tracts within 200 miles from the Fountain of Youth
4. Merge the Illudium Phosdex data to the subsetted dataset
5. Obtain a descriptive summary of the data
6. Create a histogram illustrating the new age distribution
6. Test for association between age, distance, and illudium phosdex concentrations
7. Behavior of StatCat, deviation from analysis plan
```{r}
# 1
setwd("/Users/ScottCohen/OneDrive - University of Florida/MyDataStory")
CensusAge_df<-read.csv("data/statCat_age.csv",stringsAsFactors = F)
str(CensusAge_df) 

# 2
CensusTract1990_df <-   merge(CensusAge_df, census_df[ , c("geoid","distance")], 
                        by = "geoid")

# 3
CensusTract1990_df <-   subset(CensusTract1990_df, distance < 200, 
                        select = c(geoid, county, distance, age_statcat))

# 4
PhosdexCensusTract1990_df <-    merge(CensusTract1990_df, phosdex_df[ , c("geoid","date_sampled","phosdex")], 
                                by = "geoid")

# 5
summary(PhosdexCensusTract1990_df)
str(PhosdexCensusTract1990_df)
hist(PhosdexCensusTract1990_df$age_statcat, 
     xlab="Median Age Distribution"
     )

# 6
cor.test(PhosdexCensusTract1990_df$age_statcat, PhosdexCensusTract1990_df$phosdex)
cor.test(PhosdexCensusTract1990_df$age_statcat, PhosdexCensusTract1990_df$distance)
```
